Disease Knowledge Transfer across Neurodegenerative Diseases

### 2019-03-07 paper/paper.tex
cutting down on pages

\author{R\u{a}zvan V. Marinescu\inst{1} \and Marco Lorenzi\inst{1,4} \and Stefano Blumberg\inst{1} \and 
 Alexandra L. Young\inst{1} \and Pere P. Morell\inst{1} \and Neil P. Oxtoby\inst{1} \and Arman Eshaghi\inst{1,2} \and Keir X. Yong\inst{3} \and Sebastian J. Crutch\inst{3} \and Daniel C. Alexander\inst{1}, for 
the  Alzheimer’s  Disease  Neuroimaging Initiative\thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu). As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report. A complete listing of ADNI investigators can be found at: http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}

@@

@@

\subsection{Parameter Estimation}

\newcommand{\uu}{^{(u)}}
\newcommand{\um}{^{(u-1)}}

We estimate the model parameters using a two-stage approach. In the first stage, we perform belief propagation within each agnostic unit and then within each disease model. In the second stage we jointly optimise across all agnostic units and disease models using loopy belief propagation. An overview of the algorithm is given in Figure \ref{fig:dktAlgo}. Given the initial parameters estimated from the first stage (line 1), the algorithm continuously updates the biomarker trajectories within the agnostic units (lines 4-5), dysfunction trajectories (line 8) and subject-specific time shifts (line 10) until convergence. The cost function for all parameters is nearly identical, the main difference being the measurements $(i,j,k)$ over subjects $i$, visits $j$ and biomarkers $k$ that are selected for computing the measurement error. For estimating the trajectory of biomarker $k$ within agnostic unit $\psi(k)$, measurements are taken from $\Omega_k$ representing all measurements of biomarker $k$ from all subjects and visits. For estimating the dysfunction trajectories, $\Omega_{d,l}$ represents the measurement indices from all subjects with disease $d$ (i.e. $d_i = d$) and all biomarkers $k$ that belong to agnostic unit $l$ (i.e. $\psi(k) = l$). Finally, $\Omega_i$ (line 10) represents all measurements from subject $i$, for all biomarkers and visits.

\begin{figure}
\begin{algorithm}[H]
\tiny
 Initialise $\boldsymbol{\theta}^{(0)}$, $\boldsymbol{\lambda}^{(0)}$, $\boldsymbol{\beta}^{(0)}$\\
  \While{$\boldsymbol{\theta}$, $\boldsymbol{\lambda}$, $\boldsymbol{\beta}$ not converged}{
   \tcp*[l]{Estimate biomarker trajectories (disease agnostic)}
    \For{$k=1$ to $K$}{
      ${\theta_k\uu = \argmin_{\theta_k} \sum_{(i,j) \in \Omega_k} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\right) \right]^2  - log\ p(\theta_k)}$\\
      ${\epsilon_k\uu = \frac{1}{|\Omega_k|} \sum_{(i,j) \in \Omega_k}    \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\uu \right) \right]^2 }$\\
    }
     \tcp*[l]{Estimate dysfunction trajectories (disease specific)} 
    \For{$d=1 \in \mathbb{D}$}{
      \For{$l=1 \in \Lambda$}{
        ${\lambda_{d}^{l, (u)} = \argmin_{\lambda_{d}^{l}} \sum_{(i,j,k) \in \Omega_{d,l}} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lambda_{d}^{l}) ; \theta_k\uu 
        \right) \right]^2  - log\ p(\lambda_{d}^{l})}$\\
      }
    }
    \tcp*[l]{Estimate subject-specific time shifts} 
    \For{$i=1 \in [1, \dots, S]$}{
      ${\beta_i\uu = \argmin_{\beta_i} \sum_{(j,k) \in \Omega_i} \left[y_{ijk} - g\left(f(\beta_i + m_{ij}; \lpuu) ; \theta_k\uu
      \right) \right]^2  - log\ p(\beta_i)}$\\
    }
}
\normalfont
\end{algorithm}
\caption[The algorithm for estimating the DKT parameters]{The algorithm used to estimate the DKT parameters, based on loopy belief-propagation.}
\label{fig:dktAlgo}
\end{figure}

@@

 \begin{table}
 %\fontsize{7}{12}
 \centering
 \begin{tabular}{c | c}
& \textbf{Trajectory parameters} \\
%\hline
 Biomarker allocation &  $l_0:\{k_0, k_2, k_4\}$, $l_1: \{k_1, k_3, k_5\}$ \\
Agnostic unit $l_0$ &  $\theta_0 = (1,5,0.2,0)$, $\theta_2 = (1,5,0.55,0)$,  $\theta_4 = (1,5,0.9,0)$  \\
Agnostic unit $l_1$ & $\theta_1 = (1,10,0.2,0)$, $\theta_3 = (1,10,0.55,0)$, $\theta_5 = (1,10,0.9,0)$ \\

@@

We trained DKT on ADNI data from the TADPOLE challenge \cite{marinescu2018tadpole}, since it contained a large number of multimodal biomarkers already pre-processed and aggregated into one table. From the TADPOLE dataset we selected a subset of 230 subjects which had at least one FDG PET, AV45, AV1451 or DTI scan. Most subjects also had MRI scans and cognitive tests. In order to model another disease, we further included 76 PCA subjects from the DRC centre in the training set, along with 67 tAD and 87 age-matched controls, all of which only had MRI scans.

@@

\begin{table}
\footnotesize
\begin{tabular}{c | c c c c c c}
\textbf{Model} & \textbf{Cingulate} & \textbf{Frontal} & \textbf{Hippo.} & \textbf{Occip.} & \textbf{Parietal} & \textbf{Temporal}\\
& \multicolumn{6}{c}{\textbf{Prediction Error (MSE)}}\\
DKT & 0.09$\pm$0.04 & 0.03$\pm$0.01 & 0.18$\pm$0.03 & 0.04$\pm$0.02 & 0.06$\pm$0.02 & 0.04$\pm$0.02\\
Latent stage & 0.09$\pm$0.04 & 0.03$\pm$0.01 & 0.17$\pm$0.03 & 0.04$\pm$0.02 & 0.06$\pm$0.02 & 0.04$\pm$0.02\\
Linear & 0.05$\pm$0.02* & 0.15$\pm$0.04* & 0.09$\pm$0.03* & 0.07$\pm$0.03* & 0.07$\pm$0.02* & 0.07$\pm$0.02*\\
& \multicolumn{6}{c}{\textbf{Rank Correlation (Spearman rho)}}\\
DKT  & 0.76  & 0.48  & 0.76  & 0.55  & 0.55  & 0.33 \\
Latent stage & 0.76  & 0.49  & 0.80*  & 0.56  & 0.51*  & 0.33 \\
Linear & 0.48*  & 0.31*  & 0.64*  & 0.61*  & 0.57*  & 0.27* \\
\end{tabular}
\normalfont
\caption[Performance evaluation of DKT and two simpler models]{Performance evaluation of DKT and two simpler models. (*) Statistically significant difference between the current model performance and DKT, based on a two-tailed t-test, Bonferroni corrected.}
\label{sec:dktPerfMetrics}
\end{table}

@@

While we provided an example implementation of DKT using specific models of the biomarker trajectories, measurement noise and link function (the disease progression score), DKT should be considered as a general framework for joint modelling of biomarker trajectories within different diseases simultaneously. The actual implementation of DKT can thus be extended to use non-parametric trajectories, or more complex link functions that estimate not only subject time-shifts but also progression speed or higher order terms.

While here we have focused on Alzheimer's variants such as tAD and PCA, DKT can also be applied to other progressive neurodegenerative diseases of non-Alzheimer's type and even the normal ageing process. While we only used imaging data, cognitive tests can also be included in the disease-specific sub-models of DKT, or even allocated in the agnostic units of the regions that are responsible for those tasks.

@@

As a key direction of further research, DKT can be used to understand novel mechanisms within a disease (e.g. AD variants), as well as across diseases (e.g. AD vs FTD). For AD variants in particular, one can study whether early biomarkers such as amyloid and tau have different dynamics compared to typical AD, and what these differences are. This can be done by choosing different a-priori allocations of biomarkers to agnostic units -- e.g. allocating amyloid and tau into one a shared agnostic unit, or otherwise directly to the disease specific models, thus making them dependent on the disease. The model with the best data fit or predictive power will indicate the most likely mechanism.

>>>>

\author{R\u{a}zvan V. Marinescu\inst{1} \and Marco Lorenzi\inst{1,4} \and Stefano Blumberg\inst{1} \and 
 Alexandra L. Young\inst{1} \and Pere P. Morell\inst{1} \and Neil P. Oxtoby\inst{1} \and Arman Eshaghi\inst{1,2} \and Keir X. Yong\inst{3} \and Sebastian J. Crutch\inst{3} \and Daniel C. Alexander\inst{1}, for 
the  Alzheimer’s  Disease  Neuroimaging Initiative
% \thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu). As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report. A complete listing of ADNI investigators can be found at: http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}
}

@@

We estimate the model parameters using a two-stage approach. In the first stage, we perform belief propagation within each agnostic unit and then within each disease model. In the second stage we jointly optimise across all agnostic units and disease models using loopy belief propagation.

@@

% \subsection{Parameter Estimation}
% 
% \newcommand{\uu}{^{(u)}}
% \newcommand{\um}{^{(u-1)}}
% 
% We estimate the model parameters using a two-stage approach. In the first stage, we perform belief propagation within each agnostic unit and then within each disease model. In the second stage we jointly optimise across all agnostic units and disease models using loopy belief propagation. An overview of the algorithm is given in Figure \ref{fig:dktAlgo}. Given the initial parameters estimated from the first stage (line 1), the algorithm continuously updates the biomarker trajectories within the agnostic units (lines 4-5), dysfunction trajectories (line 8) and subject-specific time shifts (line 10) until convergence. The cost function for all parameters is nearly identical, the main difference being the measurements $(i,j,k)$ over subjects $i$, visits $j$ and biomarkers $k$ that are selected for computing the measurement error. For estimating the trajectory of biomarker $k$ within agnostic unit $\psi(k)$, measurements are taken from $\Omega_k$ representing all measurements of biomarker $k$ from all subjects and visits. For estimating the dysfunction trajectories, $\Omega_{d,l}$ represents the measurement indices from all subjects with disease $d$ (i.e. $d_i = d$) and all biomarkers $k$ that belong to agnostic unit $l$ (i.e. $\psi(k) = l$). Finally, $\Omega_i$ (line 10) represents all measurements from subject $i$, for all biomarkers and visits. 
% 
% 
% \begin{figure}
% \begin{algorithm}[H]
% \tiny
%  Initialise $\boldsymbol{\theta}^{(0)}$, $\boldsymbol{\lambda}^{(0)}$, $\boldsymbol{\beta}^{(0)}$\\
%   \While{$\boldsymbol{\theta}$, $\boldsymbol{\lambda}$, $\boldsymbol{\beta}$ not converged}{
%    \tcp*[l]{Estimate biomarker trajectories (disease agnostic)}
%     \For{$k=1$ to $K$}{
%       ${\theta_k\uu = \argmin_{\theta_k} \sum_{(i,j) \in \Omega_k} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\right) \right]^2  - log\ p(\theta_k)}$\\
%       ${\epsilon_k\uu = \frac{1}{|\Omega_k|} \sum_{(i,j) \in \Omega_k}    \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\uu \right) \right]^2 }$\\
%     }
%      \tcp*[l]{Estimate dysfunction trajectories (disease specific)} 
%     \For{$d=1 \in \mathbb{D}$}{
%       \For{$l=1 \in \Lambda$}{
%         ${\lambda_{d}^{l, (u)} = \argmin_{\lambda_{d}^{l}} \sum_{(i,j,k) \in \Omega_{d,l}} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lambda_{d}^{l}) ; \theta_k\uu 
%         \right) \right]^2  - log\ p(\lambda_{d}^{l})}$\\
%       }
%     }
%     \tcp*[l]{Estimate subject-specific time shifts} 
%     \For{$i=1 \in [1, \dots, S]$}{
%       ${\beta_i\uu = \argmin_{\beta_i} \sum_{(j,k) \in \Omega_i} \left[y_{ijk} - g\left(f(\beta_i + m_{ij}; \lpuu) ; \theta_k\uu
%       \right) \right]^2  - log\ p(\beta_i)}$\\
%     }
% }
% \normalfont
% \end{algorithm}
% \caption[The algorithm for estimating the DKT parameters]{The algorithm used to estimate the DKT parameters, based on loopy belief-propagation.}
% \label{fig:dktAlgo}
% \end{figure}

@@

 \begin{table}
\fontsize{6}{8}\selectfont
 \centering
 \begin{tabular}{c | c}
& \textbf{Trajectory parameters} \\
%\hline
 Biomarker allocation &  $l_0:\{k_0, k_2, k_4\}$, $l_1: \{k_1, k_3, k_5\}$ \\
Agnostic unit $l_0$ &  $\theta_0 = (1,5,0.2,0)$, $\theta_2 = (1,5,0.55,0)$,  $\theta_4 = (1,5,0.9,0)$  \\
Agnostic unit $l_1$ & $\theta_1 = (1,10,0.2,0)$, $\theta_3 = (1,10,0.55,0)$, $\theta_5 = (1,10,0.9,0)$ \\

@@

We trained DKT on ADNI data from the TADPOLE challenge \cite{marinescu2018tadpole}, since it contained a large number of multimodal biomarkers already pre-processed and aggregated into one table. From the TADPOLE dataset we selected a subset of 230 subjects which had at least one FDG PET, AV45, AV1451 or DTI scan. Most subjects also had MRI scans and cognitive tests. In order to model another disease, we further included 76 PCA subjects from the DRC centre in the training set, along with 67 tAD and 87 age- matched controls, all of which only had MRI scans.

@@

\begin{table}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}{c | c c c c c c}
\textbf{Model} & \textbf{Cingulate} & \textbf{Frontal} & \textbf{Hippo.} & \textbf{Occip.} & \textbf{Parietal} & \textbf{Temporal}\\
& \multicolumn{6}{c}{\textbf{Prediction Error (MSE)}}\\
DKT & 0.09$\pm$0.04 & 0.03$\pm$0.01 & 0.18$\pm$0.03 & 0.04$\pm$0.02 & 0.06$\pm$0.02 & 0.04$\pm$0.02\\
Latent stage & 0.09$\pm$0.04 & 0.03$\pm$0.01 & 0.17$\pm$0.03 & 0.04$\pm$0.02 & 0.06$\pm$0.02 & 0.04$\pm$0.02\\
Linear & 0.05$\pm$0.02* & 0.15$\pm$0.04* & 0.09$\pm$0.03* & 0.07$\pm$0.03* & 0.07$\pm$0.02* & 0.07$\pm$0.02*\\
& \multicolumn{6}{c}{\textbf{Rank Correlation (Spearman rho)}}\\
DKT  & 0.76  & 0.48  & 0.76  & 0.55  & 0.55  & 0.33 \\
Latent stage & 0.76  & 0.49  & 0.80*  & 0.56  & 0.51*  & 0.33 \\
Linear & 0.48*  & 0.31*  & 0.64*  & 0.61*  & 0.57*  & 0.27* \\
\end{tabular}
% \normalfont
\caption[Performance evaluation of DKT and two simpler models]{Performance evaluation of DKT and two simpler models. (*) Statistically significant difference between the current model performance and DKT, based on a two-tailed t-test, Bonferroni corrected.}
% \end{footnotesize}
\label{sec:dktPerfMetrics}
\end{table}

@@

% While we provided an example implementation of DKT using specific models of the biomarker trajectories, measurement noise and link function (the disease progression score), DKT should be considered as a general framework for joint modelling of biomarker trajectories within different diseases simultaneously. The actual implementation of DKT can thus be extended to use non-parametric trajectories, or more complex link functions that estimate not only subject time-shifts but also progression speed or higher order terms.
% 
% While here we have focused on Alzheimer's variants such as tAD and PCA, DKT can also be applied to other progressive neurodegenerative diseases of non-Alzheimer's type and even the normal ageing process. While we only used imaging data, cognitive tests can also be included in the disease-specific sub-models of DKT, or even allocated in the agnostic units of the regions that are responsible for those tasks.

@@

% As a key direction of further research, DKT can be used to understand novel mechanisms within a disease (e.g. AD variants), as well as across diseases (e.g. AD vs FTD). For AD variants in particular, one can study whether early biomarkers such as amyloid and tau have different dynamics compared to typical AD, and what these differences are. This can be done by choosing different a-priori allocations of biomarkers to agnostic units -- e.g. allocating amyloid and tau into one a shared agnostic unit, or otherwise directly to the disease specific models, thus making them dependent on the disease. The model with the best data fit or predictive power will indicate the most likely mechanism.
